---
title: "BS 1515 Lecture 1: Sets, Inequalities, and Equations"
author: "Sergey V. Popov"
date: "`r Sys.Date()`"

output: slidy_presentation
css: cardiff.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Economics

We as economists:

- Describe what people do with allocation of resources (empirics)

- Developing stories about why they behave (theory)

Both of these require math:

- why data methods applied to data produce meaningful results (empirics)

- why assumptions about individual behavior lead to predictions (theory)

Math is how economists argue.

## Math

Math studies logic applied to things.

- Agents behaving optimally: What's an agent? How do they choose? What's the difference between optimal and not optimal? How do agents interact?
- Things being random: what things? If one thing happened, can another thing happen at the same time? If one thing happened, do we know if another might happen more likely?
- Countries developing: what resources they use? how do they combine? why different countries use similar resources with drastically different results?

Things we study today will apply to other parts of this course, and will be useful for other courses as well.

## Example: Liberation Day

April 2, 2025: Trump imposed tariffs on all other countries

- minimum 10%
- more at countries with trade deficit in goods but not services
- planning to abolish income tax, use tariff income instead

Questions:

- How is it supposed to work?
- Need to think about foreign producers and consumers
- Need to think about domestic producers and consumers
- Where does the tax formula come from

Need to argue precisely! (or cause markets to collapse)

## Sets

The fundamental of calculus is a *set*, containing *elements*.

- Set of students $\{\text{John},\text{Megan},\text{Dylan},\text{Susie}...\}$.
- Set of students in this classroom.
- Set of students who is supposed in this classroom.

An object can be *in* a set ($a \in A$) or *not in* ($a \not \in A$).

If all elements of set $B$ are in set $A$, $B$ is a *subset* of $A$: $B \subseteq A$; $A$ is a *superset* of $B$.

If there are elements in $A$ that are not in $B$, $B$ is a *proper subset* of $A$: $B \subset A$, and $A$ is a *proper superset* of $B$.

*Empty set* containing no elements is $\emptyset=\{\}$. *Singleton* set contains one element.

## Example

When shopping, people have *budgets* (say £100/week) and decide on a shopping basket

- set of all shopping baskets

When choosing where to study, pupils choose

- the set of university to apply - rules restrict
- school to attend

When deciding on the stock/bond portfolio

- proportions of assets
- amount of money to invest

## Operations on sets

We can manipulate sets:

- A *complement* of $A$ denoted as $\neg A= \{a: a \not \in A\}$
- *union* of $A$ and $B$ denoted as $A \cup B=\{a \in A: a \in B\}$. 
- *intersection* of $A$ and $B$ denoted as $A \cap B= \{a\in A: a \in B\}$. 
- *subtraction* of $B$ from $A$ denoted as $A \setminus B=\{a \in A : a \not \in B\}$.

$\land$ means AND. $\lor$ means OR. $\Rightarrow$ means part on the left implies part on the right. $\Leftrightarrow$ means when left part holds, right part holds, and vice versa.

Prove: $A \setminus B = A \cap \neg B$.

Prove: $A \cap B \subseteq A \cup B$. When can we claim that $A \cup B \subset A \cap B$?

Prove: $A \cup B = (A \cap B) \cup (A \setminus B) \cup (B \setminus A)$.

## Euler-Venn diagram

```{r,engine='tikz',fig.alt="Two overlapping sets to illustrate the operations with sets"}
\begin{tikzpicture}
 \draw[fill=red!40!white, draw=black, fill opacity=0.3] (0,0) circle (0.6cm) node[yshift=-1cm, color=red!40!white, fill opacity=1] {$A$};
 \draw[fill=green!40!black, draw=black, fill opacity=0.3] (1,0) circle (0.6cm) node[yshift=-1cm, color=green!40!black, fill opacity=1] {$B$};
\end{tikzpicture}
```

*Space* is important!

## Exercises for Sets

Consider a 52-cards standard deck. A poker hand is 5 cards drawn from this deck at random. Verify claims:

- The set of poker hands featuring three of a kind is a subset of a set of poker hands featuring a pair.

- The set of poker hands featuring two pairs is a subset of a set of poker hands featuring a pair.

- The set of poker hands featuring a straight (cards in sequence of rank ignoring suites) and the set of poker hands featuring a pair do not overlap.

- The set of poker hands featuring a flush (5 cards of the same suit) and the set of poker hands containing a straight do not overlap.

- The set of poker hands containing full house is an intersection of hands containing three of a kind and hands that contain a pair.

## Functions

*Function* $f:A\rightarrow B$ is a rule that gives an element of $B$ for every element of $A$.

$f(x)=2x+3$ is an example of $f:\mathbb{R}\rightarrow \mathbb{R}$.

$F(x,y)=0$ is an *implicit function*.

- Not always a "true" function that gives you $y$ of $x$!
- $F(x,y)=y-2x-3=0$ gives you a function $y=2x+3$.
- $F(x,y)=y^2-x=0$ gives you two values of $y$ for each value of $x\geq 0$.
- We'll come back to these later.

*One-to-one* function: for each element of $A$ you get a *unique* element of $B$.

*Onto* function: for each $b\in B$ there is an  $a\in A$ that gives $f(a)=b$.

If both, function is *invertible*. $f(x)=x^2$ is invertible when defined on $\mathbb{R}_+\rightarrow \mathbb{R}_+$ but not if defined on $\mathbb{R}\rightarrow \mathbb{R}$.

## Examples of Functions

- $f(x)=x^2$ is invertible for $x \in [0,+\infty)$ but not for $x\in(-\infty,+\infty)$.

- $f(x)=e^x$ is invertible, $g(y)=\ln(y)$ is the inverse.

- $f(x)=a+bx$ is invertible if $b\not =0$.

Why invertibility matters?

- If we have a function from unobservables to data, we want to infer unobservables from data

## Real numbers
*Naturals* $\mathbb{N}=\{1,2,3,..\}$

*Integers* $\mathbb{Z}=\mathbb{N}\cup\{0\}\cup(-\mathbb{N})$

*Rationals* $\mathbb{Q}=\{a/b: a, b\in \mathbb{N}\}$

*Reals* $\mathbb{R}$ are different: includes $\mathbb{Q}$ and *irrationals* such as $\sqrt{2}$. Reals have a property that  $\forall X,Y \text{ such that }\forall x \in X \text{ and }\forall y \in Y x<y \Rightarrow \exists z:x<z<y$.

Notice that $\mathbb{Q}$ does not have this property: $X=\{t:t^2<2\}$ and $Y=\{t:t^2>2\}$ are not separated by any $z \in \mathbb{Q}$!

*Euclidean space* $\mathbb{R}^N$ for $N\in \mathbb{N}$.

*Metric*: way of measuring distance in space, a function $d(x,y)$. For reals, could be $d(x,y)=|x-y|$. Metric is a feature of the space.

Metrics must be 

(a) positive: $d(x,y)\geq 0$;

(b) symmetric: $d(x,y)=d(y,x)$; and

(c) $d(x,y)\leq d(x,z)+d(z,y)$ - triangular inequality.

## Important Reals

$\pi=3.14159...$ is not rational.

$e=2.7182828...$ is not rational.

Neither can be represented as a solution to a polynomial equation with integer coefficients, or *transcendental*.

## Countable Sets

Set is *countable* if there is an invertible function from the set to $\mathbb{N}$.

Set is *finite* if there is no onto function from $\mathbb{N}$ to the set.

Set is *uncountable* otherwise.

Prove: $\mathbb{Q}$ is countable. (hard!)

Prove: $\mathbb{R}$ is uncountable. (hard!)

## Open and Closed

Open $\varepsilon$-ball around $x$: $B_\varepsilon(x)=\{t:d(x,t)<\varepsilon\}$.

A set $A$ is *open* in space *S* with metric $d(x,y)$ if $\forall x \in A \exists \varepsilon>0: B_{\varepsilon}(x) \subset A$.

A set $A$ is *closed* if complement of $A$, $\neg A=S\setminus A$, is open.

Prove: A union of open sets is open.

Prove: An intersection of closed sets is closed.

Prove: A space $S$ is both closed and open in $S$.

Prove: An empty set $\emptyset$ is both closed and open in $S$.


## Order and Intervals

Some spaces are *ordered*: there is a function $R(x,y)$ such that $R(x,y)=TRUE$ or $R(x,y)=FALSE$ for $x,y \in S$.

- $\mathbb{R}$ is ordered with respect to ``bigger than'' order $>$
- $\mathbb{R^2}$ is partially ordered with respect to $>(x,y)=TRUE$ if $x_1>y_1\land x_2>y_2$.
- If you have a *utility function* over your *consumption bundles*, your preference relation is ordered.

Interval $(a,b) \subseteq S$ is $\{x:a<x<b\}$. Interval $[a,b] \subseteq S$ is $\{x:a\leq x \leq b\}$.

Prove: Interval $(a,b)$ in $\mathbb{R}$ is open.

Prove: Interval $[a,b]$ in $\mathbb{R}$ is closed.

Prove: Interval $(a,b]$ in $\mathbb{R}$ is neither open nor closed.

## Convex set

Take two points in a set $A$, $a$ and $b$. Take $c(\lambda)=\lambda a+(1-\lambda)b$ for $\lambda\in(0,1)$.

- Can you always do this? What do you need to take a point like this?

If for every $a,b, \lambda$ point $c(\lambda)$ is in set $A$, this set is *convex*.

Prove: an intersection of two convex sets is convex.

Prove: empty set is convex.

Prove: any single element set is convex.

Prove: the space set is convex.

## Equalities and Inequalities

Consider a set

\[\left\{(x,y)\in\mathbb{R}^2:x+2y<3\right\}\]

Plot it:

```{r,engine='tikz',fig.alt="Illustrating an inequality"}
\usetikzlibrary{patterns}
\centering \begin{tikzpicture}[x=1cm,y=1cm]
\draw[->] (-4.1,0)--(4.1,0) node[right] {$x$};
\draw[->] (0,-2.1)--(0,2.1) node[right] {$y$};
\foreach \x in {-4,...,-1,1,...,4}
\draw[] (\x,0.1)--(\x,-0.1) node[below] {$\x$};
\foreach \x in {-2,...,-1,1,...,2}
\draw[] (-0.1,\x)--(0.1,\x) node[right] {$\x$};
\begin{scope}
\clip (-4,-2) rectangle (4,2);
 \draw[pattern=north west lines, pattern color=red!40!white, draw=red!40!white, thick, dashed] (-6,4.5)--(6,-1.5)--(6,-6)--(-6,-6);
 \end{scope}
\end{tikzpicture}
```

Is that an open set? Now do the same with equality.

## System of Inequalities

Now we have two inequalities:

\[\left\{(x,y)\in\mathbb{R}^2:x+2y<3,x+y>2\right\}\]

Plot it:

```{r,engine='tikz',fig.alt="Illustrating two inequalities in two dimensional space"}
\usetikzlibrary{patterns}
\centering \begin{tikzpicture}[x=1cm,y=1cm]
\draw[->] (-4.1,0)--(4.1,0) node[right] {$x$};
\draw[->] (0,-2.1)--(0,2.1) node[right] {$y$};
\foreach \x in {-4,...,-1,1,...,4}
\draw[] (\x,0.1)--(\x,-0.1) node[below] {$\x$};
\foreach \x in {-2,...,-1,1,...,2}
\draw[] (-0.1,\x)--(0.1,\x) node[right] {$\x$};
\begin{scope}
\clip (-4,-2) rectangle (4,2);
 \draw[pattern=north west lines, pattern color=red!40!white, draw=red!40!white, thick, dashed] (-6,4.5)--(6,-1.5)--(6,-6)--(-6,-6);
 \draw[pattern=north east lines, pattern color=green!40!white, draw=green!40!white, thick, dashed] (-3,5)--(5,-3)--(4,4);
 \end{scope}
\end{tikzpicture}
```

Is that an open set? 

## Equation

Consider a set

\[\left\{(x,y)\in\mathbb{R}^2:x+2y=3\right\}\]

Plot it:

```{r,engine='tikz',fig.alt="Illustrating an equation"}
\usetikzlibrary{patterns}
\centering \begin{tikzpicture}[x=1cm,y=1cm]
\draw[->] (-4.1,0)--(4.1,0) node[right] {$x$};
\draw[->] (0,-2.1)--(0,2.1) node[right] {$y$};
\foreach \x in {-4,...,-1,1,...,4}
\draw[] (\x,0.1)--(\x,-0.1) node[below] {$\x$};
\foreach \x in {-2,...,-1,1,...,2}
\draw[] (-0.1,\x)--(0.1,\x) node[right] {$\x$};
\begin{scope}
\clip (-4,-2) rectangle (4,2);
 \draw[draw=red!40!white, thick, dashed] (-6,4.5)--(6,-1.5)--(6,-6)--(-6,-6);
 \end{scope}
\end{tikzpicture}
```

Is that an open set?

## System of Equations

Now we have two equations:

\[\left\{(x,y)\in\mathbb{R}^2:x+2y=3,x+y=2\right\}\]

Plot it:

```{r,engine='tikz',fig.alt="Illustrating two equations in two dimensional space"}
\usetikzlibrary{patterns}
\centering \begin{tikzpicture}[x=1cm,y=1cm]
\draw[->] (-4.1,0)--(4.1,0) node[right] {$x$};
\draw[->] (0,-2.1)--(0,2.1) node[right] {$y$};
\foreach \x in {-4,...,-1,1,...,4}
\draw[] (\x,0.1)--(\x,-0.1) node[below] {$\x$};
\foreach \x in {-2,...,-1,1,...,2}
\draw[] (-0.1,\x)--(0.1,\x) node[right] {$\x$};
\begin{scope}
\clip (-4,-2) rectangle (4,2);
 \draw[draw=red!40!white, thick, dashed] (-6,4.5)--(6,-1.5)--(6,-6)--(-6,-6);
 \draw[draw=green!40!white, thick, dashed] (-3,5)--(5,-3)--(4,4);
 \end{scope}
\end{tikzpicture}
```

Generically - do you have to have one intersection? What if equations are not linear?

## Example: Supply and Demand

Let Supply be $Q=3p-20$ and Demand be $Q=60-p$. What is the equilibrium price?

Answer 1: $3p-20=60-p$ so $4p=80$ and $p=20$ and $Q=60-20=40$.

Answer 2: solve a system of $Q=60-p$ and $Q=3p-20$.

Now you know: how do we know there is a solution?..

## Continuous Functions

A function $f:D\rightarrow E$ is *continuous* at $x$ if for every $\varepsilon>0$ there is a $\delta>0$ such that for all $x'\in B_\varepsilon(x)$ it holds that $f(x') \in B_\delta (f(x))$.

*Intermediary Value Theorem*: For continuous function $f(\cdot)$, if $f(a)=f_1$ and $f(b)=f_2$, there is $x$ in $[a,b]$ such that $f(x)=f$ for every $f$ in $[f_1,f_2]$.

Important part: $f(\cdot)$ is *continuous*: no jumps!

Automatic for linear functions, but not generic.

Good news for us: can establish that an equation/system of equations has a solution, but not necessarily uniqueness.

## Maximum Theorem

Let $f(x,y)$ be continuous function on a compact set. Then $max_x f(x,y)$ exists and is continuous in $y$.

Extends to multiple dimensions and constraints.

*Compact* means contains all limits. Closed and bounded in $\mathbb{R}^N$, but not enough for $\mathbb{R}^\infty$!

Good news for us: people maximizing utility will choose something as long as their constraints make sense.

## Plan for next week

We will talk more about derivatives and how do we use them for optimization. We will use the notions of functions and sets from this lecture.